PAMELA: an annotation-based Java Modeling Framework

### 2021-02-16 6-Experimentations.tex
typos, harmonization of vocabulary, remove colors, ...

Model Federation~\cite{Golra2016} is an approach that provides the means to
integrate multiple models conforming to different paradigms, and giving to each
stakeholder a specific view adapted to its needs. Model federation approach is
developed as a possible response to SIMF RFP (Semantic Information Modeling for
Federation)i~\cite{simf-rfp} by OMG (Object Modeling Group). This RFP (Request For Proposal)
requests submissions for a standard addressing, ”federation of information
across different representations, levels of abstraction, communities,
organizations, viewpoints, and authorities”. Thus model federation allows the
integration of heterogeneous models to develop new cross-concern
viewpoints/models or to synchronize the models used for designing a system.

@@

\sg{
The full Openflexo infrastructure is composed of about 50 components. In most components, PAMELA framework is extensively used. As an example \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagramming features) is composed of 998 classes. 159 of those classes (mostly the diagramming model) are defined as Pamela \texttt{ModelEntity}. 
}

@@

\sg{
An interesting experimentation has been done in the context of Openflexo development process. When Pamela has been integrated to the code base, a big portion of former legacy code has gradually and iteratively been migrated to Pamela. Refactoring mainly consisted in removing code, and replacing method implementation by API method declaration. In some parts of core model implementation, code has been reduced by 80\% (in terms of lines of Java code), and many bugs disappeared, as they were caused by programming errors.
}

\sg{
Pamela implementation is now really stable and mature. According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no Pamela-specific issue, while co-evolution of model and code is greatly improved compared to a "code generation"-based solution.
}

>>>>

Model Federation~\cite{Golra2016} is an approach that provides the means to
integrate multiple models conforming to different paradigms, and giving to each
stakeholder a specific view adapted to its needs. Model federation approach is
developed as a possible response to SIMF RFP (Semantic Information Modeling for
Federation)~\cite{simf-rfp} by OMG (Object Modeling Group). This RFP (Request For Proposal)
requests submissions for a standard addressing, ”federation of information
across different representations, levels of abstraction, communities,
organizations, viewpoints, and authorities”. Thus model federation allows the
integration of heterogeneous models to develop new cross-concern
viewpoints/models or to synchronize the models used for designing a system.

@@

The full Openflexo infrastructure is composed of about 50 components. In most components, PAMELA framework is extensively used. As an example \texttt{diana} component\footnote{\url{https://diana.openflexo.org}} (a component providing diagramming features) is composed of 998 classes. 159 of those classes (mostly the diagramming model) are defined as Pamela \texttt{ModelEntity}.

@@

An interesting experimentation has been done in the context of Openflexo development process. When Pamela has been integrated to the code base, a big portion of former legacy code has gradually and iteratively been migrated to Pamela. Refactoring mainly consisted in removing code, and replacing method implementation by API method declaration. In some parts of core model implementation, code has been reduced by 80\% (in terms of lines of Java code), and many bugs disappeared, as they were caused by programming errors.

Pamela implementation is now really stable and mature. According to Openflexo infrastructure developers, maintenance of code base (about 10 years of development) raises no Pamela-specific issue, while co-evolution of model and code is greatly improved compared to a "code generation"-based solution.

### 2021-02-16 architecture.tex
typos, harmonization of vocabulary, remove colors, ...

A \texttt{PAMELAModel} is defined as a set of references to \texttt{ModelEntity}. A \texttt{ModelEntity} reflects a concept and is encoded in a Java \texttt{interface}. PAMELA metamodel allows multiple inheritance: thus \texttt{ModelEntity} may define a set of parent entities. Note that reification of \texttt{ModelEntity} is performed in a Java \texttt{interface} (and not a class), which only defines API whithout any implementation for methods. 
Note that \emph{ModelEntities} may declare \emph{ImplementationClasses}, which
will be responsible for providing custom code implementing domain or
application specific behaviour. A partially implemented \texttt{abstract} Java
\texttt{class} may be defined as partial base implementation (conforms to
implemented interface), \jcb{as default methods in Java interfaces would be
(since Java 8).}

@@

% Folloging section/itemize may be removed when needed 
%This composition offers many benefits: 
%\begin{itemize}
%    \item Strong coupling between model and code
%    \item Strong typing is kept, and required checks are performed by the Java compiler
%    \item PAMELA framework provides interpretation of model@runtime
%    \item No need to generate POJO (plain old Java objects), as their
%    execution follow the standard semantics (less code, fewer bugs)
%    \item Custom implementation are provided if needed, using classical Java extension points
%    \item It offers a way to intercept method calls and instrument the code
%    \item Assertions checking at runtime
%    \item Dynamic code weaving at runtime (aspect programming without compilation)
%\end{itemize}

>>>>

A \texttt{PAMELAModel} is defined as a set of references to \texttt{ModelEntity}. A \texttt{ModelEntity} reflects a concept and is encoded in a Java \texttt{interface}. PAMELA metamodel allows multiple inheritance: thus \texttt{ModelEntity} may define a set of parent entities. Note that reification of \texttt{ModelEntity} is performed in a Java \texttt{interface} (and not a class), which only defines API whithout any implementation for methods. 
Note that \emph{ModelEntities} may declare \emph{ImplementationClasses}, which
will be responsible for providing custom code implementing domain or
application specific behaviour. A partially implemented \texttt{abstract} Java
\texttt{class} may be defined as partial base implementation (conforms to
implemented interface), as default methods in Java interfaces would be
(since Java 8).

@@

% Folloging section/itemize may be removed when needed 
%This composition offers many benefits: 
%\begin{itemize}
%    \item Strong coupling between model and code
%    \item Strong typing is kept, and required checks are performed by the Java compiler
%    \item PAMELA framework provides interpretation of model@runtime
%    \item No need to generate POJO (Plain Old Java Objects), as their
%    execution follow the standard semantics (less code, fewer bugs)
%    \item Custom implementation are provided if needed, using classical Java extension points
%    \item It offers a way to intercept method calls and instrument the code
%    \item Assertions checking at runtime
%    \item Dynamic code weaving at runtime (aspect programming without compilation)
%\end{itemize}

### 2021-02-16 implementation.tex
typos, harmonization of vocabulary, remove colors, ...

The PAMELA framework is a 100\% pure Java ($\geq$ /1.81.8), compilable by a classical Java compiler and executable in a classical Java Virtual Machine.

@@

\sg{

@@

}
%\todo{Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}

@@

\sg{
Compared to a basic POJO (Plain Object Java Object) implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpretated nature for some model features.

@@

}

>>>>

The PAMELA framework is a 100\% pure Java ($\geq$1.8), compilable by a classical Java compiler and executable in a classical Java Virtual Machine.

@@

@@

%\todo{Surely you can say more about its structure, size of components, ... other interesting and relevant observations on the codebase?}

@@

Compared to a basic POJO (Plain Old Java Object) implementation, use of PAMELA involves a significant CPU and memory-footprint overhead, because of the partially-interpretated nature for some model features.

@@

### 2021-02-16 pamelaProcess.tex
typos, harmonization of vocabulary, remove colors, ...

We advocate for a strong coupling between model and source-code, to give architects and developers a way to both interact during the whole development cycle. PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization. \sg{Instead of generating the code, the API (mostly Java interfaces with abstract method declarations) is locally executed (interpreted).} The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues. Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture and is more detailed thereafter.

@@

\sg{
Here are various conceivable scenarios for PAMELA use:
\begin{itemize}
    \vspace{-0.2cm}\item \textbf{Programming use} : programmers reuse existing annotations and write model and code at the same time (the modelers and programmers share the same artifacts : the Java code). This scenario includes the case where the model (made by the modelers) is pre-existing.
    \vspace{-0.2cm}\item \textbf{Reengineering} : programmers start from an existing base of code (legacy) and refactor it while replacing this code by abstract method declarations in Java interface, reducing risks of errors.
    \vspace{-0.2cm}\item \textbf{Aspect-oriented programming} : programmers may use or redefine "patterns" (e.g., Security Patterns \cite{silva20}) which offers code weaving at runtime, and runtime monitoring.
    \vspace{-0.2cm}\item \textbf{Advanced programmers use} : programmers may extend PAMELA with their own annotations, implementations or patterns.
\end{itemize}
}

>>>>

We advocate for a strong coupling between model and source-code, to give architects and developers a way to both interact during the whole development cycle. PAMELA is an annotation-based Java modeling framework providing a smooth integration between model and code, without code generation nor externalized model serialization. Instead of generating the code, the API (mostly Java interfaces with abstract method declarations) is locally executed (interpreted). The idea is to avoid separation between modeling and code to facilitate consistency management and to /avoid round-trip issues. Figure \ref{fig:PamelaVision} summarizes the PAMELA architecture and is more detailed thereafter.

@@

Here are various conceivable scenarios for PAMELA use:
\begin{itemize}
    \vspace{-0.2cm}\item \textbf{Programming use} : programmers reuse existing annotations and write model and code at the same time (the modelers and programmers share the same artifacts : the Java code). This scenario includes the case where the model (made by the modelers) is pre-existing.
    \vspace{-0.2cm}\item \textbf{Reengineering} : programmers start from an existing base of code (legacy) and refactor it while replacing this code by abstract method declarations in Java interface, reducing risks of errors.
    \vspace{-0.2cm}\item \textbf{Aspect-oriented programming} : programmers may use or redefine "patterns" (e.g., Security Patterns \cite{silva20}) which offers code weaving at runtime, and runtime monitoring.
    \vspace{-0.2cm}\item \textbf{Advanced programmers use} : programmers may extend PAMELA with their own annotations, implementations or patterns.
\end{itemize}

### 2021-02-16 related.tex
typos, harmonization of vocabulary, remove colors, ...

The PAMELA framework can be seen as a CASE tool that focuses in the design and verification and validation phases of the software development life-cycle. In that sense, it present similarities to other existing object oriented CASE tools and/or language workbenches such as Kermeta~\cite{kermeta} and the Platypus~\cite{platypus} meta-case tool as they permit the definition and manipulation of models, metamodels and constraints/behavior. Two main features separate PAMELA from those approaches though: 1) The main artefact for the PAMELA framework is standard java code; and 2) PAMELA does not rely on code generation but on the interpretation/construction of models at runtime guided by code annotations. This allows for the seamlessly blending of the metamodeling/programming phases.

>>>>

The PAMELA framework can be seen as a CASE tool that focuses in the design and verification and validation phases of the software development life-cycle. In that sense, it presents similarities to other existing object oriented CASE tools and/or language workbenches such as Kermeta~\cite{kermeta} and the Platypus~\cite{platypus} meta-case tool as they permit the definition and manipulation of models, metamodels and constraints/behavior. Two main features separate PAMELA from those approaches though: 1) The main artefact for the PAMELA framework is standard java code; and 2) PAMELA does not rely on code generation but on the interpretation/construction of models at runtime guided by code annotations. This allows for the seamlessly blending of the metamodeling/programming phases.
